---
title: "GSE33000_download_full"
author: "Shradha Mukherjee"
date: "March 1, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####################################Getting raw data GSE33000####################################### 

###Step1: Load required libraries ans setting working directory###

```{r}
#save working directory location
wd<-getwd()
wd
```

```{r}
#Install packages by uncommenting two lines below if packages not already installed before
#source("https://bioconductor.org/biocLite.R")
#biocLite(c("GEOquery","Biobase", "limma", "R.utils", "lumi", "DT"))

library(Biobase)
library(GEOquery)
library(lumi)
library(R.utils)
library(DT)
library(limma)
library(sva)
library(pamr)
```

```{r}
#At start of pipeline no input files are needed except the .Rproj which sets working directory to current directory and keeps R environment independent of other folders and the .Rmd R code
list.files()
```

###Step2: Getting GEO expression data and metadata###
  #2.1 Get GE0 data as R object
```{r}
# Saving series and platform gene annotation data from GEO to gset object
gset <- getGEO("GSE33000", GSEMatrix =TRUE, AnnotGPL=TRUE)
if (length(gset) > 1) idx <- grep("GPL4372", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]] 
```

  #2.2 keep only samples of interest
```{r}




#Interested in all samples so keeping everything. 




```

  #2.3 Getting GEO metadata of interest
```{r}
#Names of available phenotype data
names(pData(gset))
```

```{r}
metadata=data.frame(gset$geo_accession, gset$source_name_ch1, gset$`tissue:ch1`, gset$`gender:ch2`, gset$`age:ch2`,gset$`disease status:ch2`)
head(metadata)
write.csv(metadata,'GSE33000_metadata.csv')
```

  #2.4 Get GE0 raw expression data. 
```{r}
DT::datatable(exprs(gset)[1:3,])
```

```{r}
#The expression data uploaded for this entry is Agilent dual color cy5 cy3 label microarray 
exprList<-exprs(gset)
write.table(exprList, "GSE33000_expression_data_originalfromGEO.txt", sep="\t", quote=F)
```

###Step3: Quantile normalization with lumi ###
  #3.1 Reading and formatting expression data for normalization
```{r}
#Now we read in the raw expression values
exprRaw<-read.delim('GSE33000_expression_data_originalfromGEO.txt',header=T, sep='\t')
DT::datatable(exprRaw[1:3,])
DT::datatable(summary(exprRaw))
```

```{r}
#The original GSE33000 is in log base 10 ratio (Cy5/Cy3)  https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1423780

#conversiton back to non-log scale
exprRaw1 = 10^exprRaw
DT::datatable(exprRaw1[1:3,])
write.table(exprRaw1, "GSE33000_expression_data_non-log.txt", sep="\t", quote=F)
```

  #3.2 Using lumi for quantile normalization  
```{r}
#Perform qualtile normalization on the raw expression data should be matrix format.
exprLumi<-lumiN(as.matrix(exprRaw1),method="quantile")
DT::datatable(exprLumi[1:3,])
```

  #3.3 log2+1 transformation
```{r}
#log2+1 transform the expression data. This step also changes it back to data frame format
exprLog<-log2(exprLumi+1)
DT::datatable(exprLog[1:3,])
```

  #3.4 Check for negative values. Negative values will not work in WGCNA
```{r}
exprRaw_ifneg<-apply(exprRaw, 1, function(row) any(row <0))
length(which(exprRaw_ifneg)) #what is the length of negative numbers

exprRaw1_ifneg<-apply(exprRaw1, 1, function(row) any(row <0))
length(which(exprRaw1_ifneg)) #what is the length of negative numbers
#0 means no negative values

exprLumi_ifneg<-apply(exprLumi, 1, function(row) any(row <0))
length(which(exprLumi_ifneg)) #what is the length of negative numbers
#0 means no negative values 

exprLog_ifneg<-apply(exprLog, 1, function(row) any(row <0))
length(which(exprLog_ifneg)) #what is the length of negative numbers
#0 means no negative values 
```

  #3.5 Check for max and min values. Ususally don't want max to be <100 
```{r}
exprRaw_max<-which.max(as.matrix(exprRaw))
exprRaw_max
exprRaw_min<-which.min(as.matrix(exprRaw))
exprRaw_min

exprRaw1_max<-which.max(as.matrix(exprRaw1))
exprRaw1_max
exprRaw1_min<-which.min(as.matrix(exprRaw1))
exprRaw1_min

exprLumi_max<-which.max(as.matrix(exprLumi))
exprLumi_max
exprLumi_min<-which.min(as.matrix(exprLumi))
exprLumi_min

exprLog_max<-which.max(as.matrix(exprLog))
exprLog_max
exprLog_min<-which.min(as.matrix(exprLog))
exprLog_min
```

  #3.6 Descriptive or summary statistics of the data
```{r}
DT::datatable(summary(exprRaw))
```

```{r}
DT::datatable(summary(exprRaw1))
```

```{r}
DT::datatable(summary(exprLumi))
```

```{r}
DT::datatable(summary(exprLog))
```

	#3.7 Visualization of GSE33000 alone
```{r}
pdf(file="Step3_plot_Visualization_GSE33000_rawdata_transformation.pdf",height=5,width=5)
par(mfrow=c(2,2))

#randomly select 16 samples to visualize change in exprssion data with normalization. Trying to visualiza all sample is too many and hard to see
col_sel=sample(ncol(exprRaw), 16) #use number less than 16 is total samples is less than 16

#exprRaw boxplot
par(mai=c(1,0.8,1,0.8))
boxplot(exprRaw[,col_sel], outline=FALSE, las=2, cex=0.25, main="exprRaw", col="yellow")

#exprRaw1 boxplot
par(mai=c(1,0.8,1,0.8))
boxplot(exprRaw1[,col_sel], outline=FALSE, las=2, cex=0.25, main="exprRaw1", col="yellow")

#exprLumi boxplot
par(mai=c(1,0.8,1,0.8))
boxplot(exprLumi[,col_sel], outline=FALSE, las=2, cex=0.25, main="exprLumi", col="yellow")

#exprLog boxplot
par(mai=c(1,0.8,1,0.8))
boxplot(exprLog[,col_sel], outline=FALSE, las=2, cex=0.25, main="exprLog", col="yellow")


#exprRaw MDSplot, colored by Study
par(mai=c(1,0.8,1,0.8))
plotMDS(exprRaw[,col_sel], legend= "all", main="exprRaw Study", cex=0.5)#like PCA plot

#exprRaw1 MDSplot, colored by Study
par(mai=c(1,0.8,1,0.8))
plotMDS(exprRaw1[,col_sel], legend= "all", main="exprRaw1 Study", cex=0.5)#like PCA plot

#exprLumi MDSplot, colored by Study
par(mai=c(1,0.8,1,0.8))
plotMDS(exprLumi[,col_sel], legend= "all", main="exprLumi Study", cex=0.5)#like PCA plot

#exprLog MDSplot, colored by Study
par(mai=c(1,0.8,1,0.8))
plotMDS(exprLog[,col_sel], legend= "all", main="exprLog Study", cex=0.5)#like PCA plot

dev.off()
```

###Step4: Use GPL file to convert probe ID to GeneSymbol in exprLog that will be used for further analysis ###
  #4.1 Getting GPL annotation file 
```{r}
GPLid<-annotation(gset)
GPL_file<-getGEO(GPLid)
## File stored at:
## /var/folders/0g/5tz57sgx7090n5vttfqvs3ym0000gn/T//Rtmpv5hBUt/GPL4372.soft
colnames(GPL_file@dataTable@table)
write.table(GPL_file@dataTable@table,"GPL4372.txt", sep='\t', quote=F)
```

```{r}
#We pick the ID or reporter ID and ORF or gene symbols from options displayed above. We need this to annotate expression data. Will use ID to combine.
GPL_file1<-GPL_file@dataTable@table[,c("ID","ORF")]
DT::datatable(head(GPL_file1))
```

  #4.2 preparing gene expression data exprLog for merging with annotation file
```{r}
exprGene=exprLog
#colnames already are sample GSM id
DT::datatable(head(exprGene))
```

```{r}
#Add "ID" column that we will use for merging with annotation
exprGene1=cbind(exprGene, rownames(exprGene))
colnames(exprGene1)=c(colnames(exprGene),"ID")
DT::datatable(head(exprGene1))
```

  #4.3 get gene expression data with gene symbols. GEO annotation is static and ensures that annotation of genes does not change over time.
```{r}
#This is complete gene expression with Gene symbol annotations
#Save and use columns for pipelines as needed
exprGene2=merge(exprGene1, GPL_file1, by="ID")
dim(exprGene2)
DT::datatable(head(exprGene2))
write.csv(exprGene2, "GSE33000_Annotation_Expr_GeneHu.csv")
```

```{r}
#The files we needed are saved so we now clear workspace and delete files and folders that are not needed 
save.image(file="preSVA_GSE33000_temp.RData")
rm(list=ls())
gc()
```
